---
title: "N741 Homework 8"
author: "Alyssa Monda"
date: "April 12, 2017"
output:
  github_document: default
  html_document: default
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

 In this text file, as you read through it and scroll down, you'll see the following:

```
7. Attribute Information: (class attribute has been moved to last column)

   #  Attribute                     Domain
   -- -----------------------------------------
   1. Sample code number            id number
   2. Clump Thickness               1 - 10
   3. Uniformity of Cell Size       1 - 10
   4. Uniformity of Cell Shape      1 - 10
   5. Marginal Adhesion             1 - 10
   6. Single Epithelial Cell Size   1 - 10
   7. Bare Nuclei                   1 - 10
   8. Bland Chromatin               1 - 10
   9. Normal Nucleoli               1 - 10
  10. Mitoses                       1 - 10
  11. Class:                        (2 for benign, 4 for malignant)
```
#Import Data
```{r}

#The raw data files can be downloaded from the associated Data Folder at [http://archive.ics.uci.edu/ml/machine-learning-databases/breast-cancer-wisconsin/](http://archive.ics.uci.edu/ml/machine-learning-databases/breast-cancer-wisconsin/). 
#CSV comma delimited file with NO column names in the 1st row. The datafile description and associated column file names are in the "breast-cancer-wisconsin.names" which is a simple text file

#Import breat-cancer-wisconsin.data
library(readr)
bcw <- read_csv("~/Desktop/Big Data/N741hmwk8/breast-cancer-wisconsin.data")
#View(bcw)

#import breat-cancer-wisconsin.names
library(readr)
bcw_names <- read_csv("~/Desktop/Big Data/N741hmwk8/breast-cancer-wisconsin.names")
#View(bcw_names)
```


The combined dataset has 699 cases (rows). However, 16 cases were missing values for the "Bare Nuclei" measurement. The R code below, processes the data, applies the names, and removes the cases with missing values. So, the final dataset created below `bcdat` will have 683 cases and 11 variables.

```{r}
# from tidyverse - use readr
# to read in the comma delimited dataset
library(readr)

# raw data does not have column names
bcdat <- read_csv("breast-cancer-wisconsin.data",
                  col_names=FALSE)

# add variable names
names(bcdat) <- c("idnum","clumpthickness","uniformcellsize",
                  "uniformcellshape","marginaladhesion",
                  "singlecellsize","barenuclei","blandchromatin",
                  "normalnucleoli","mitoses","class")

# note in column 7 "Bare Nucleoli" there are
# question marks "?" that need to be set to missing NA
library(dplyr)
bcdat <- bcdat %>%
  mutate(barenucfix = ifelse(barenuclei=="?",NA,
                              as.numeric(barenuclei)))

# keep the main 11 variables
bcdat <- bcdat %>%
  select(idnum,clumpthickness,uniformcellsize,uniformcellshape,
         marginaladhesion,singlecellsize,barenucfix,blandchromatin,  
         normalnucleoli,mitoses,class)

# keep only complete cases, n=683
bcdat <- na.omit(bcdat)
```

# Principal Components Analysis

## 1. Perform the PCA

```{r}
# use only columns 2 through 10
# you do not need the idnum, nor the class variables
pr.out <- prcomp(bcdat[,2:10], scale=TRUE)
summary(pr.out)
```

## 2. Make plots of the variance and PVE 

### Plot of the Variances of Each PC

```{r}
plot(pr.out)
```

### Plot of the PVE and Cumulative PVE of each PC

```{r}
# plots of the PVE percent variance explained
pve = 100*pr.out$sdev^2/sum(pr.out$sdev^2)
pve
plot(pve, type = "o", ylab = "Cumulative PVE", xlab = "Principal Component", col="blue")
plot(cumsum(pve), type = "o", ylab = "Cumulative PVE", xlab = "Principal Component", col="brown3")

```

## 3. Make a "loadings plot" of the variables

```{r}
# loadings are in the "rotation" part of the 
# pr.out list object. "rotation" is a matrix
# with a row for each variable and a column for
# each PC.
pr.out$rotation

# choose the 1st and 2nd columns for the 1st 2 PCs
# and plot these loading weights for the 9
# variables. I tweaked the limits some
# feel free to change these as needed
plot(pr.out$rotation[,1],pr.out$rotation[,2],
     xlim=c(-0.5,0.1),ylim=c(-0.5,1),
     cex=2, pch=19,
     xlab = "Principal Component 1",
     ylab = "Principal Component 2",
     main = "Loadings Plot for PC 1 and 2")

# add xpd=FALSE to prevent lines drawn outside plot area
par(xpd=FALSE)

# add red dashed lines for the axes at y=0 and x=0
abline(h=0, col="red")
abline(v=0, col="red")

# overlay the variable names on this loading plot
text(pr.out$rotation[,1],pr.out$rotation[,2],
     labels = rownames(pr.out$rotation),
     pos = 3)
```

## 4. Scores Plot on 1st 2 PCs

```{r}
# scores plot - use x from the pr.out list object
# plot scores on 1st 2 PCs, columns 1 and 2 of x
# color the points by the "class" variable for
# benign (class=2) or malignant (class=4)
plot(pr.out$x[,1],pr.out$x[,2], 
     col = bcdat$class,
     xlab = "Principal Component 1",
     ylab = "Principal Component 2",
     main = "Scores Plot on PC 1 and 2",
     sub = "Blue=Benign (class=2) and Red=Malignant (class=4)")
```

## Homework 8 Tasks

1. Rerun the PCA (steps 1-4 above) for (A) just the Benign cases and for just the (B) Malignant Cases. The code below, sets up these data subsets for you.

```{r}
# Benign cases
bcdatBenign <- bcdat %>%
  filter(class == 2)

# Malignant cases
bcdatMalignant <- bcdat %>%
  filter(class == 4)
```

#Run Steps for bcdatBenign


```{r}
# redo for benign ==============
bcdat <- bcdatBenign
# run steps above
```

## 1. Perform the PCA

```{r}
# use only columns 2 through 10
# you do not need the idnum, nor the class variables
pr.out <- prcomp(bcdat[,2:10], scale=TRUE)
summary(pr.out)
```

## 2. Make plots of the variance and PVE 

### Plot of the Variances of Each PC

```{r}
plot(pr.out)
```

### Plot of the PVE and Cumulative PVE of each PC

```{r}
# plots of the PVE percent variance explained
pve = 100*pr.out$sdev^2/sum(pr.out$sdev^2)
pve
plot(pve, type = "o", ylab = "Cumulative PVE", xlab = "Principal Component", col="blue")
plot(cumsum(pve), type = "o", ylab = "Cumulative PVE", xlab = "Principal Component", col="brown3")

```

## 3. Make a "loadings plot" of the variables

```{r}
# loadings are in the "rotation" part of the 
# pr.out list object. "rotation" is a matrix
# with a row for each variable and a column for
# each PC.
pr.out$rotation

# choose the 1st and 2nd columns for the 1st 2 PCs
# and plot these loading weights for the 9
# variables. I tweaked the limits some
# feel free to change these as needed
plot(pr.out$rotation[,1],pr.out$rotation[,2],
     xlim=c(-0.5,0.1),ylim=c(-0.5,1),
     cex=2, pch=19,
     xlab = "Principal Component 1",
     ylab = "Principal Component 2",
     main = "Loadings Plot for PC 1 and 2")

# add xpd=FALSE to prevent lines drawn outside plot area
par(xpd=FALSE)

# add red dashed lines for the axes at y=0 and x=0
abline(h=0, col="red")
abline(v=0, col="red")

# overlay the variable names on this loading plot
text(pr.out$rotation[,1],pr.out$rotation[,2],
     labels = rownames(pr.out$rotation),
     pos = 3)
```

## 4. Scores Plot on 1st 2 PCs

```{r}
# scores plot - use x from the pr.out list object
# plot scores on 1st 2 PCs, columns 1 and 2 of x
# color the points by the "class" variable for
# benign (class=2) or malignant (class=4)
plot(pr.out$x[,1],pr.out$x[,2], 
     col = bcdat$class,
     xlab = "Principal Component 1",
     ylab = "Principal Component 2",
     main = "Scores Plot on PC 1 and 2",
     sub = "Blue=Benign (class=2) and Red=Malignant (class=4)")
```


#Run Steps for bcdatMalignant
```{r}
# redo for malignant ==================
bcdat <- bcdatMalignant
# run steps above
```


## 1. Perform the PCA

```{r}
# use only columns 2 through 10
# you do not need the idnum, nor the class variables
pr.out <- prcomp(bcdat[,2:10], scale=TRUE)
summary(pr.out)
```

## 2. Make plots of the variance and PVE 

### Plot of the Variances of Each PC

```{r}
plot(pr.out)
```

### Plot of the PVE and Cumulative PVE of each PC

```{r}
# plots of the PVE percent variance explained
pve = 100*pr.out$sdev^2/sum(pr.out$sdev^2)
pve
plot(pve, type = "o", ylab = "Cumulative PVE", xlab = "Principal Component", col="blue")
plot(cumsum(pve), type = "o", ylab = "Cumulative PVE", xlab = "Principal Component", col="brown3")

```

## 3. Make a "loadings plot" of the variables

```{r}
# loadings are in the "rotation" part of the 
# pr.out list object. "rotation" is a matrix
# with a row for each variable and a column for
# each PC.
pr.out$rotation

# choose the 1st and 2nd columns for the 1st 2 PCs
# and plot these loading weights for the 9
# variables. I tweaked the limits some
# feel free to change these as needed
plot(pr.out$rotation[,1],pr.out$rotation[,2],
     xlim=c(-0.5,0.1),ylim=c(-0.5,1),
     cex=2, pch=19,
     xlab = "Principal Component 1",
     ylab = "Principal Component 2",
     main = "Loadings Plot for PC 1 and 2")

# add xpd=FALSE to prevent lines drawn outside plot area
par(xpd=FALSE)

# add red dashed lines for the axes at y=0 and x=0
abline(h=0, col="red")
abline(v=0, col="red")

# overlay the variable names on this loading plot
text(pr.out$rotation[,1],pr.out$rotation[,2],
     labels = rownames(pr.out$rotation),
     pos = 3)
```

## 4. Scores Plot on 1st 2 PCs

```{r}
# scores plot - use x from the pr.out list object
# plot scores on 1st 2 PCs, columns 1 and 2 of x
# color the points by the "class" variable for
# benign (class=2) or malignant (class=4)
plot(pr.out$x[,1],pr.out$x[,2], 
     col = bcdat$class,
     xlab = "Principal Component 1",
     ylab = "Principal Component 2",
     main = "Scores Plot on PC 1 and 2",
     sub = "Blue=Benign (class=2) and Red=Malignant (class=4)")
```



2. In the overall dataset, when looking at the loadings plot, which variables cluster together? which variables do not lie with that cluster?
-Looking at the loading plot for the dataset as a whole, it appears everything clusters together except mitoses. Mitoses has a higher loading on component 1 and 2 than the other variables. 


3. How do the variable clusters seen in the loading plots for the -Benign data subset and Malignant subset differ? and how are they similar if at all? 
-For the benign subset it appears less of the variables cluster. Single cell size, normal nuclei, uniform cell share, and uniform cell size cluster. Barencfix and marginal adhesision cluster together. And clump thickness and bland chromatin cluster together. Mitoses appears again here with higher loading on both components, far from any other variables. 
-For the malignant subset mitoses appears to cluster with a few other variables, at least with normal nuclei and single cell size. uniform cell size and unifrom cell shape appear to form a separate cluster. Bland chromatin and marginal adhesision cluster together. While barenucfix shows high loading on component 1 separating it from the other variables, and clump thickness shows high loading on component 1 and 2. 
-There do not appear to be obvious similarities between the two plots. Uniform cell size and uniform cell shape appear to cluster together on both plots, but otherwise there are more differences recognizeable on the plots than similarities. Mainly the differences in loading value of mitoses clump thickness and barenucfix between the 2 plots. 


4. Is using 2 principal components reasonable for summarizing the variability seen in this Breast Cancer dataset with 9 measurements? Explain your reasoning for (a) the overall dataset, (b) the Benign subset and (c) the Malignant subset
(a) Using 2 principal components is reasonable for summarizing the variability seen in the data set as a while. The bar plot of the variances for the entire data set shows most of the variance is found within component 1 so using both component 1 and 2 to summarize the variability seems reasonable considering the low variances described by the other components. 
(b) For the benign data set component 1 still represents the majority of the variance. While the variances explained by the other components appears to increase, component 1 still represents over twice as much variance compared to the other components. Because component 2 and 3 appear to represent similar variance it may be worth investigating if adding a 3rd dimension to the plot to see if any new relationships appear. However I would say 2 components was still reasonable for summariing the variability. 
(c) For the malignant data set component 1 continues to show much of the variance however, the variance described by components 2 and 3 are much higher. In this case it may be beneficial to have a thrid axis to represent the variability in component 3, at least to investigate if it changes the clustering or displays a relationship that is not seen in 2-D. 

5. While PCA is an unsupervised data analysis method (i.e. no "target" class information is used in the analysis), do you think the 2 PCs extracted do a good job of helping to distinguish Benign cases from Malignant cases (i.e. look back at the overall dataset Scores Plot). Explain your rationale.

-Based on the overall scores plots I do believe the 2 PCs did a good job distinguishing the benign and malignant cases. The "red" and "blue" clusters are fairly distinguishable, as the malignant cases are more highly loaded on component 1. Based on the bar plot of the variances most of the variance in the dataset was found in component 1, so plotted in the overall dataset scores plot the clustering is distinguishable on the axis for component 1 (as seen in the plot).


6. This markdown file used to create this document can be found in the repository at <https://github.com/amonda/N741hmwk8.git> and is names N741Hmwk8.Rmd. 